---
title: "PS5"
author: "Lily Bates"
date: "2023-02-14"
output: html_document
---

# Problem Set 5: Gapminder data

## 1 Load and check data (5pt)
You first task is to do a very simple data check:
1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown docu-
ment with an appropriate title. See https://faculty.washington.edu/otoomet/info201-book/
r-markdown.html#r-markdown-rstudio-creating.


```{r message=FALSE}
library(tidyverse)
library(readr)

gapminder <- read_delim("gapminder.csv.bz2")
```

2. (2pt) Load data. How many rows/columns do we have?

```{r}
dim(gapminder)
```


3. (2pt) Print a small sample of data. Does it look OK?

```{r}

gapminder %>% 
  sample_n(3)

```

This data looks good! 

2 Descriptive statistics (15pt)
1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

```{r}
num_names <- gapminder %>% 
  summarize(n = n_distinct(name))

num_iso3 <- gapminder %>% 
  summarize(n = n_distinct(iso3))

num_iso2 <- gapminder %>% 
  summarize(n = n_distinct(iso2))
```

In the gapminder dataset , there are `r num_names` names, `r num_iso3` 3-digit
country codes, and `r num_iso2` 2-digit country codes. 

### 2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?
(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?

 

```{r}
gapminder %>% 
  filter(is.na(iso2), time == 2019) %>% 
  select(name)
```

There are 149 iso2 codes for 150 countries, so one country does not have an iso2
code, that country is Namibia. Using time 2019 is arbitary, I used it as a 
placeholder so all observations were not inlcuded in the table.

(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.

```{r}
gapminder %>% 
  filter(is.na(name), time == 2019) %>% 
  select(iso3)

```

Yes,m the CHANISL, GBM, KOS, and NLD CURACAO have more than one iso3 code for 
the country they repersent.

3. (2pt) What is the minimum and maximum year in these data?

```{r}
maxyear<- gapminder %>% 
  filter(!is.na(time)) %>% 
  summarise(max(time))

minyear<- gapminder %>% 
  filter(!is.na(time)) %>% 
  summarise(min(time))
```

The minimum year in this data is `r minyear` and maximum year is `r maxyear`

## 3 CO2 emissions (30pt)

Next, let’s analyze CO2 emissions.
1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?

```{r}
gapminder %>% 
  group_by(time) %>% 
  summarise(missingCO2 = sum(is.na(co2)),
            missingCO2_PC = sum(is.na(co2_PC))) %>% 
  arrange(desc(missingCO2))
  
```

The year 2017, 2018, and 2019 had the most missing data with 2017 missing data
enteries

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.

```{r}
plottingData <- gapminder %>% 
  filter(!is.na(time) , !is.na(co2), !is.na(co2_PC)) %>% 
  filter(iso3 =="USA" | iso3 =="IND"| iso3 =="CHN"| iso3 =="SWE"|iso3 =="COL")
  
ggplot(data = plottingData) +
  geom_point(aes(time, co2, color=iso3))+
  labs(x= "year", y= "CO2 Emissions (kt)")

```

The US and China are the two largest CO2 emission countries, with China's 
emissions spiking over the US in the last 20 years. Sweden and Colombia are
consistently the lowest CO2 emsission countries. 

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the
same countries. What does this figure suggest?

```{r}
ggplot(data = plottingData) +
  geom_point(aes(time, co2_PC, color=iso3))+
  labs(x="year",y="CO2 Emissions Per Capita (tons)")

```

USA residents have the highest per capita CO2 emissions throughout the data
and Swedens citizens are second. Indian citizens are the lowest but have 
seen an increase in the last 50 years. 

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different
size.

```{r}
gapminder %>%
  filter(!is.na(co2_PC),
         !is.na(region)) %>% 
  filter(time %in% c(1960,2016)) %>% 
  group_by(time, region) %>%
  summarize(aveCO2PC = mean(co2_PC))
```
In 1960, America is the top producer of CO2 per capita, followed by Europe, 
Oceania, Asia, and Africa. 

In 2016, Europe is the top producer of CO2 per capita, followed by Asia, 
Americas, Oceania, and Africa. 

5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.

```{r}

gapminder %>%
  filter(!is.na(co2_PC),
         !is.na(region)) %>% 
  filter(time== 1960 | time == 2016) %>% 
  group_by(region, time) %>%
  mutate(aveCO2PC = mean(co2_PC))%>% 
ggplot(aes(region, aveCO2PC, fill = factor(time))) +
  geom_col(stat = 'identity', position = 'dodge')+
  labs(x= "Continent", y="CO2 Emissions Per Capita (ton)")
```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent).

### Americas

```{r}

gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Americas") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  head(3)

```
```{r}
gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Americas") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  tail(3)
```

### Asia

```{r}

gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Asia") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  head(3)
 

```
```{r}
gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Asia") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  tail(3)
```

### Europe

```{r}

gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Europe") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  head(3)
 

```
```{r}
gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Europe") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  tail(3)
```

### Africa

```{r}

gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Africa") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  head(3)
 

```
```{r}
gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Africa") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  tail(3)
```

### Oceania

```{r}

gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Oceania") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  head(3)
 

```
```{r}
gapminder %>% 
  filter(time == 2016, !is.na(co2_PC), region=="Oceania") %>%  
  group_by(name) %>% 
  arrange(rank(desc(co2_PC))) %>% 
  select(name, region, co2_PC) %>% 
  tail(3)
```

## 4 GDP per capita (50pt)

Let’s look at GDP per capita (GDP_PC ).
1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.

```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(GDP_PC), !is.na(region), time == 1960) %>% 
  ggplot(aes(lifeExpectancy, GDP_PC, color = region, size= totalPopulation))+
  geom_point() +
  labs(x= "Life Expectancy", y= "GDP per capita")
```

LifeE is positively associated with GDP per capita, with Europe and the America's
trending to have the higher lifeE and GDP per Capita, and Africa having the lowest. 

2. (4pt) Make a similar plot, but this time use 2019 data only.

```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(GDP_PC),!is.na(region), time == 2019) %>% 
  ggplot(aes(lifeExpectancy, GDP_PC, color = region, size= totalPopulation))+
  geom_point() +
  labs(x= "Life Expectancy", y= "GDP per capita")

```

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

The world has become larger, richer, and older in the last 60 years. Overall Life
E has moved to the right, or increased, and GDP per capita has increased in almost 
every country(this is understated in the graph due to the y-axis scale changing, 
if the data is considered with the same y-axis scale, it has moved far more up/ 
countries have gotten richer.)

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.

```{r}
gapminder %>%
  filter(!is.na(lifeExpectancy),
         !is.na(region)) %>% 
  filter(time %in% c(1960,2016)) %>% 
  group_by(time, region) %>%
  summarize(aveLifeE = mean(lifeExpectancy))
```
Yes, this result matched the graphs in that all of the regions saw increases
in lifeE! 

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.

```{r}
gapminder %>% 
  mutate(prev = lag(lifeExpectancy), 
         growth= lifeExpectancy - prev) %>% 
  filter(!is.na(growth), !is.na(region)) %>% 
  group_by(region) %>%
  summarise(aveLEGrowth = mean(growth)) %>% 
  arrange(rank(desc(aveLEGrowth)))
```
This table shows that Europe and Oceania have seen the highest growth in LifeE, 
while Africa has seen the smallest. This is consistent with the above graphs, 
as African countries, noted in red, have not moved as far to the right as
European and Asian countries between 1960 and 2019. 

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!

```{r}
gapminder %>%
  filter(!is.na(GDP_PC),
         !is.na(region)) %>% 
  filter(time== 1960 | time == 2016) %>% 
  group_by(time) %>%
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(binwidth=5000, position = "dodge") +
  labs(x= "Average GDP per capita")
  
```

*Fun fact! That outlier with nearly $200,000 GDP per capita is the small nation
of Monaco*


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.

```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time == 1960) %>% 
  mutate(LErank = rank(desc(lifeExpectancy))) %>% 
  select(iso3, name, LErank) %>% 
  filter(iso3== "USA")

```

```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time == 2019) %>% 
  mutate(LErank = rank(desc(lifeExpectancy))) %>% 
  select(iso3, name, LErank) %>% 
  filter(iso3== "USA")

```

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?

```{r}
rows1960 <- gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time == 1960) %>% 
  nrow()

gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time == 1960) %>% 
  mutate(LErank = rank(desc(lifeExpectancy)),
         LERR = LErank/rows1960) %>% 
  select(name, iso3, LErank, LERR) %>% 
  filter(iso3 == "USA")

```

```{r}

rows2019 <- gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time == 2019) %>% 
  nrow()

gapminder %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time == 2019) %>% 
  mutate(LErank = rank(desc(lifeExpectancy)),
         LERR = LErank/rows2019) %>% 
  select(name, iso3, LErank, LERR) %>% 
  filter(iso3 == "USA")

```

Finally tell us how many hours did you spend on this PS.

About 6

